---
title: "Using S4 methods for type checking"
output:
html_document: default
html_notebook: default
---

R is a weekly typed language.  This can cause issues for programmers using libraries by not learning
of issues with calls until the call is made or recieving cryptic messags when an invalid type is passed
to a function and somewhere in the bowls of said function that invalid type causes a problem

Basic first argument type checking can be done with the S3 object system but the S4 object system 
provides multi-argument type checking.  Since S4 uses Generic Functions it is possible to use the S4
system for type checking.

**Refresher**: The S3 and S4 object systems separate the "object" part, the part with the data, from the
methods.  So, it is possible to leverage the S4/S5 method dispatching without defining a new object type.  
Therefore it is possible to use the S3 and S4 method dispatching to perform type checking on a function
call without defnining a whole new object.

First, lets show what happens when a function is called with invalid parameters.

```{r error=TRUE}
myFn = function(name, sex, age, ...) {
  sprintf("%s is a %s who is %d years old.\n", name, sex, age) -> s
  cat(s)
}

myFn("Mark", "M", 50)
myFn(1, 4.3, "Spoon")
```

The first call worked fine, no issues.  The second, howevver had a problem because we passed in numbers
for characters and characters for numbers.  The error message, though, is for the sprintf function 
which is in the function we called.  Showing the Stack trace

```
Error in sprintf("%s is a %s who is %d years old.\n", name, sex, age) : invalid format '%d'; use format %s for character objects
2.
sprintf("%s is a %s who is %d years old.\n", name, sex, age)
1.
myFn(1, 4.3, "Spoon")

```

We see that myFn was called and then sprintf and then the error from sprintf.  It would be much nicer if
the error were at the myFn, since we passed the incorrect values into it in the first place.  It would
be even more helpful if it happened before the program ran so if this call were late in a long running
program we find out right away instead of after the program runs for 2 cups of coffee and an
uncomfortable conversation with the overly talkative cube mate.  Unfortunately you'll have to go to
a strongly typed language for that, like Haskel.

However, we can emulate stronger type checking using the _setGeneric()_ and _setMethod()_ functions.  Let's give it a try.

```{r error=TRUE}
myFn = NULL

setGeneric('myFn',
           def=function(name, sex, age, ...) {
             standardGeneric('myFn')
           })

setMethod('myFn',
          signature=signature(name='character',
                              sex='character',
                              age='numeric'),
          definition=function(name, sex, age, ...) {
            sprintf("%s is a cool %s who is %d years old.\n", name, sex, age) -> s
            cat(s)
          })

myFn("Mark", "M", 50)
myFn("Fred", "M", 100)
myFn(1, 4.3, "Spoon")
```

This is a bit better, we can see from the output that the problem is with the call to myFn and that it
wasn't able to find a signature for "numeric", "numeric", "character", vs. the less obvious sprintf 
error.

But, it's a bit more complicated to define the function.  Let's step through the parts to see what
each one is doing.

```
myFn = NULL
```

First off, we need to make sure there aren't any possible previous definitions of _myFn()_.  This is because
if R can't find a match for our function, it will fall back to the default version.  This is probably not
what we want because then we get the sprintf error instead of the actual problem.

We could have also `rm()` (removed) the function name, but if the function doesn't exist, _rm()_ will 
error out, setting it to NULL doesn't do that.

```
setGeneric('myFn',
def=function(name, sex, age, ...) {
standardGeneric('myFn')
})
```

The S4 object system uses Generic functions like the S3.  We need to register our generic function.  The
_def_ named argument indicates the function to call when we see the _myFn()_ function called.  The default
action is to call _standardGeneric()_ which dispatches the S4 function, passing it the name to dispatch.

```
setMethod('myFn',
signature=signature(name='character',
sex='character',
age='numeric'),
definition=function(name, sex, age, ...) {
sprintf("%s is a cool %s who is %d years old.", name, sex, age) -> s
cat(s)
})
```

Next we call the _setMethod()_ function.  This creates a "Generic" S4 function.  The first argument is the
name of the function as it will be called.  Next we specify the signature, this is what does the type
checking.  When the S4 object system sees the function _myFn_ it looks for a version which has 
three arguments of 'character', 'character', 'numeric' types in that order.  If it doesn't find one, then
it complains.  The definition is the actual function.  You can either define it with an anonymous function
as here or define a function and assign it here.  The defined function must have a different name, that is
it can't be _myFn()_.  Also, defining a function and assigning it here means that function could be called
directly if someone knows about it.  So, it's safest to define it as an anonymous function in the 
_setMethod()_ call.

# Citations
Most of the ideas were borrowed from

Modeling Data With Functional Programming in R
by Brian Lee Yung Rowe
web site: https://cartesianfaith.com/
